Prospects of detecting gamma-ray emission from galaxy clusters: cosmic rays and dark matter annihilations

### 2011-05-13 paper.tex
minor changes

In indirect detection, on the other hand, in particular in the photon
channel, one searches for products of DM annihilation in the galactic
halo and beyond. Particularly interesting targets are the galactic
center (for a recent possible indication of a signal, see
\cite{2010arXiv1010.2752H}, however, see also \cite{2010arXiv1012.5839B}), the dwarf galaxies surrounding the MW
\cite{Strigari:2006rd,Essig:2009jx,2010JCAP...01..031S}, galaxy
clusters (the focus of this work)
\cite{Ghigna:1998vn,Lewis:2002mfa,Boyarsky:2006kc,2006A&A...455...21C,2009PhRvL.103r1302P},
and even the cosmological large scale structure
\cite{Bergstrom:2001jj,Ullio:2002pj,Taylor:2002zd,Elsaesser:2004ap,Cuoco:2010jb,Abazajian:2010sq,Abdo:2010dk,Zavala:2011tt}.

@@

The electrons injected from annihilating DM also suffer from diffusive
and radiative losses. Hence we have to calculate the equilibrium
spectrum of the electrons plus positrons denoted by $\dd n_\e/\dd \ee$
in Eq.~(\ref{eq:ICemiss}). We derive this stationary solution using
the CR transport equation (neglecting convection and re-acceleration
effects):
\begin{eqnarray}
\label{eq:CRdiffusion}
\frac{\partial}{\partial t}\left(\frac{\dd n_\e}{\dd \ee}\right) = 
&\nabla& \left[D\left(\ee,\bx\right)\nabla\frac{\dd n_\e}{\dd \ee}\right] + 
\frac{\partial}{\partial \ee}
\left[b\left(\ee,\bx\right) \frac{\dd n_\e}{\dd \ee}\right]
 \nonumber \\
&+& q_\e(\ee,\bx)\,,
\end{eqnarray}
where $D(\ee,\bx)$ denotes the diffusion coefficient and $b(\ee,\bx)$
the energy loss term. The source function $q_\e(\ee,\bx)$ yields the
number of electrons and positrons produced per unit time, energy and
volume element at the position $\bx$:
\begin{equation}
q_\e(\ee,r)=\sum_f\frac{\dd N_{\e,f}}{\dd \ee}(\ee) B_f \Gamma_f(r) \,,
\end{equation}
where the annihilation rate density $\Gamma_f(r)$ is defined in
Eq.~(\ref{eq:ann_rate}). The sum runs over the kinematically allowed
annihilation final states $f$, each with a branching ratio $B_f$ and a
differential spectrum $\dd N_\e^f/\dd \ee$ that represents the
number of electrons plus positrons resulting from an annihilation
event. For neutralinos annihilating directly into electrons and
positrons we model the spectral distribution with $\dd
  N_\e^f/\dd \ee= 2\delta(\ee-m_\chi c^2)$. We use {\sc DarkSUSY} to
compute both the spectra when neutralinos annihilate directly into a
$\mu^+$ and $\mu^-$ in the LP model as well as in the four BM models
where a fraction of the annihilating neutralinos is converted into
electrons and positrons (see Sect.~\ref{sect:PF} and
Fig.~\ref{fig:q_DM} for further details).
%(see \cite{2009JCAP...01..016B} and references therein,
%also see fit).

The electrons and positrons loose their energy on a timescale that is
shorter than the diffusive timescale in the ICM of galaxy clusters for
CR electrons which is larger than the Hubble time in our energy range
\cite{1997ApJ...487..529B,2011A&A...527A..99E}. Hence, we neglect the
first term of the r.h.s. in Eq.~(\ref{eq:CRdiffusion}), and derive an
expression for the equilibrium number density:
\begin{eqnarray}
{\frac{\dd n_\e}{\dd \ee}}\left(\ee, r \right) &=&
 \frac{1}{b(\ee, r)} \int_{\ee}^{\mx c^2} \dd \ee' \, 
  q_\e(\ee', r),
\label{eq:nds}\\
b(\ee,r) &=& \tilde{b}
\left[\frac{B^2_{\rm CMB}}{8\pi}+\frac{B^2(r)}{8\pi}+u_\sd(r)\right] \ee^2\,,
\\\tilde{b}&=&\frac{4\sigma_{\rm T}c}{3(m_{\rm e}c^2)^2}\,.
\end{eqnarray}
Here we include the three main radiative loss processes for the CR
electrons and positrons: (1) IC losses on CMB photons with the
equivalent field strength of the CMB of $B_{\rm
  CMB}=3.24\mu\rmn{G}\,(1+z)^2$, where $z$ is the cosmological
redshift. (2) Synchrotron losses on ambient magnetic fields where we
parameterize the magnetic field in the galaxy cluster by $B(r) =
3\mu\rmn{G} \,[n_{\rm e}(r)/n_{\rm e}(0)]^{\alpha_\B}$. We adopt a
magnetic decline of $\alpha_\B=0.7$ in this work which follows from
flux frozen magnetic fields. (3) IC losses on starlight and dust with
an energy density $u_\sd(r)$ given by Eq.~(\ref{eq:u_SD_r}), which we
derive in the following.

>>>>

In indirect detection, on the other hand, in particular in the photon
channel, one searches for products of DM annihilation in the galactic
halo and beyond. Particularly interesting targets are the galactic
center (for a recent possible indication of a signal, see
\cite{2010arXiv1010.2752H}, however, see also \cite{2010arXiv1012.5839B}), the dwarf galaxies surrounding the MW
\cite{Strigari:2006rd,Essig:2009jx,2010JCAP...01..031S}, galaxy
clusters (the focus of this work)
\cite{Ghigna:1998vn,Lewis:2002mfa,Boyarsky:2006kc,2006A&A...455...21C,2009PhRvL.103r1302P},
and even the cosmological large scale structure
\cite{Bergstrom:2001jj,Ullio:2002pj,Taylor:2002zd,Elsaesser:2004ap,2011MNRAS.tmp..503C,Abazajian:2010sq,Abdo:2010dk,Zavala:2011tt}.

@@

The electrons injected from annihilating DM also suffer from diffusive
and radiative losses. Hence we have to calculate the equilibrium
spectrum of the electrons plus positrons denoted by $\dd n_\e/\dd \ee$
in Eq.~(\ref{eq:ICemiss}). We derive this stationary solution using
the cosmic ray transport equation (neglecting convection and
re-acceleration effects):
\begin{eqnarray}
\label{eq:CRdiffusion}
\frac{\partial}{\partial t}\left(\frac{\dd n_\e}{\dd \ee}\right) = 
&\nabla& \left[D\left(\ee,\bx\right)\nabla\frac{\dd n_\e}{\dd \ee}\right] + 
\frac{\partial}{\partial \ee}
\left[b\left(\ee,\bx\right) \frac{\dd n_\e}{\dd \ee}\right]
 \nonumber \\
&+& q_\e(\ee,\bx)\,,
\end{eqnarray}
where $D(\ee,\bx)$ denotes the diffusion coefficient and $b(\ee,\bx)$
the energy loss term. The source function $q_\e(\ee,\bx)$ yields the
number of electrons and positrons produced per unit time, energy and
volume element at the position $\bx$:
\begin{equation}
q_\e(\ee,r)=\sum_f\frac{\dd N_{\e,f}}{\dd \ee}(\ee) B_f \Gamma_f(r) \,,
\end{equation}
where the annihilation rate density $\Gamma_f(r)$ is defined in
Eq.~(\ref{eq:ann_rate}). The sum runs over the kinematically allowed
annihilation final states $f$, each with a branching ratio $B_f$ and a
differential spectrum $\dd N_\e^f/\dd \ee$ that represents the
number of electrons plus positrons resulting from an annihilation
event. For neutralinos annihilating directly into electrons and
positrons we model the spectral distribution with $\dd
  N_\e^f/\dd \ee= 2\delta(\ee-m_\chi c^2)$. We use {\sc DarkSUSY} to
compute both the spectra when neutralinos annihilate directly into a
$\mu^+$ and $\mu^-$ in the LP model as well as in the four BM models
where a fraction of the annihilating neutralinos is converted into
electrons and positrons (see Sect.~\ref{sect:PF} and
Fig.~\ref{fig:q_DM} for further details).
%(see \cite{2009JCAP...01..016B} and references therein,
%also see fit).

The electrons and positrons loose their energy on a timescale that is
shorter than the diffusive timescale in the ICM of galaxy clusters for
cosmic ray electrons which is larger than the Hubble time in our
energy range \cite{1997ApJ...487..529B,2011A&A...527A..99E}. Hence, we
neglect the first term of the r.h.s. in Eq.~(\ref{eq:CRdiffusion}),
and derive an expression for the equilibrium number density:
\begin{eqnarray}
{\frac{\dd n_\e}{\dd \ee}}\left(\ee, r \right) &=&
 \frac{1}{b(\ee, r)} \int_{\ee}^{\mx c^2} \dd \ee' \, 
  q_\e(\ee', r),
\label{eq:nds}\\
b(\ee,r) &=& \tilde{b}
\left[\frac{B^2_{\rm CMB}}{8\pi}+\frac{B^2(r)}{8\pi}+u_\sd(r)\right] \ee^2\,,
\\\tilde{b}&=&\frac{4\sigma_{\rm T}c}{3(m_{\rm e}c^2)^2}\,.
\end{eqnarray}
Here we include the three main radiative loss processes for the cosmic
ray electrons and positrons: (1) IC losses on CMB photons with the
equivalent field strength of the CMB of $B_{\rm
  CMB}=3.24\mu\rmn{G}\,(1+z)^2$, where $z$ is the cosmological
redshift. (2) Synchrotron losses on ambient magnetic fields where we
parameterize the magnetic field in the galaxy cluster by $B(r) =
3\mu\rmn{G} \,[n_{\rm e}(r)/n_{\rm e}(0)]^{\alpha_\B}$. We adopt a
magnetic decline of $\alpha_\B=0.7$ in this work which follows from
flux frozen magnetic fields. (3) IC losses on starlight and dust with
an energy density $u_\sd(r)$ given by Eq.~(\ref{eq:u_SD_r}), which we
derive in the following.
